Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-05-03 manuscript/manuscript.qmd
Fix targets workflow

---
title: "Cognate Priming"
subtitle: "Lab notes"
date: "Updated: `r format(Sys.Date(), '%Y/%m/%d')`"
authors:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    corresponding: true
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias fargas 25-27
        city: Barcelona
        state: Spain
        postal-code: 08005
abstract: "Previous studies have provided evidence that lexical access is language non-selective in bilinguals: recognising and producing words in one language activates lexical representations of words in the other language (e.g., Costa et al., 2000; Thierry & Wu, 2007). It has been suggested that this parallel activation is already present during toddlerhood (Jardak & Byers-Heinlein, 2019; Von Holzen & Mani, 2012), but it is unclear how it impacts the developing lexicon. In the present study, we tested Catalan-Spanish and English-Spanish bilinguals in a word recognition task, in which they were presented with words in their dominant language exclusively. In each trial, two pictures were presented side-by-side and one of them was named (target picture). Participants' visual preference for the named picture was taken as an indicator of word recognition. Each pair of pictures was preceded by the presentation of a silent image (prime picture). We designed three types of trials: (1) cognate trials in which prime and target labels shared phonological onset and the prime label was a cognate (e.g., flower-flor / fork), (2) non-cognate trials, in which prime and target labels shared phonological onset but the prime label was a non-cognate (e.g., frog-rana / fork), and (3) unrelated trials, in which prime and target labels did not share onset (unrelated trials; car-coche / fork). In line with previous studies, we expected participants to generate implicit labels for the prime pictures, which should interfere with target word recognition when both words share phonological onset (e.g., Mani & Plunkett, 2011). Additionally, under the hypothesis that bilingual participants would activate labels for the prime pictures in both languages, we predicted that interference would be stronger after cognate primes (labels from both languages share phonological overlap with target word recognition) than after non-cognate primes (only the label in the target language share phonological overlap with target word recognition). We tested bilinguals (n = 46) and same-aged monolingual controls, n = 123) at three age points (21, 25, and 30 months) to investigate how any cross-language priming effect emerged or changed across these ages."
format:
    html:
        link-external-newwindow: true
        number-offset: 0
        code-fold: true
        code-overflow: scroll
        code-line-numbers: true
        code-copy: hover
        reference-location: margin
    pdf: default
    docx:
        reference-doc: reference.docx
fig-dpi: 1000
warning: false
echo: false
toc: true
colorlinks: true
number-sections: true
bibliography: "references.bib"
csl: "apa7.csl"
---

@@

@@

@@

```{r}
#| label: tbl-participants-lp
#| tbl-cap: "Description of language profile of participants. Data are summarised for each age group, and for monolinguals and bilinguals separately."
participants %>% 
	select(id, age_group, age, lp, doe_catalan, doe_spanish, test_language) %>% 
	mutate(id = paste0(id, " (", age_group, ")")) %>%
	add_count(lp, name = "n_lp") %>% 
	add_count(age_group, name = "n_age_group") %>%

	pivot_longer(starts_with("doe_"), names_to = "language", values_to = "doe") %>%
	add_count(age_group, test_language, name = "n_age_test") %>% 
	mutate(language = str_remove_all(language, "doe_") %>% 
		   	str_to_sentence(),
		   age_group = paste0(age_group, " (N = ", n_age_group, ")"),
		   test_language = paste0("Tested in ", test_language, " (N = ", n_age_test, ")"),
		   lp = factor(lp, levels = rev(unique(lp))))  %>% 
	group_by(age_group, lp, test_language, language) %>% 
	summarise(across(doe, lst(mean, sd)),
			  .groups = "drop") %>% 
	pivot_wider(id_cols = c(age_group, test_language),
				names_from = c(language, lp),
				values_from = c(matches("doe"))) %>% 
	clean_names() %>% 
	relocate(age_group, test_language,
			 matches("catalan"),
			 matches("spanish")) %>% 
	gt(rowname_col = "test_language", 
	   groupname_col = "age_group", row_group.sep = ": ") %>% 
	tab_spanner(md("Monolingual (*N* = 162)"), matches("monolingual")) %>% 
	tab_spanner(md("Bilingual (*N* = 133)"), matches("bilingual")) %>%
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) %>% 
	cols_merge_uncert(col_val = doe_mean_catalan_monolingual,
					  col_uncert = doe_sd_catalan_monolingual) %>% 
	cols_merge_uncert(col_val = doe_mean_catalan_bilingual,
					  col_uncert = doe_sd_catalan_bilingual) %>% 
	cols_merge_uncert(col_val = doe_mean_spanish_monolingual, 
					  col_uncert = doe_sd_spanish_monolingual) %>% 
	cols_merge_uncert(col_val = doe_mean_spanish_bilingual, 
					  col_uncert = doe_sd_spanish_bilingual) %>%

	cols_label(doe_mean_catalan_monolingual = "Catalan (%)",
			   doe_mean_catalan_bilingual = "Catalan (%)",
			   doe_mean_spanish_monolingual = "Spanish (%)",
			   doe_mean_spanish_bilingual = "Spanish (%)") %>% 
	tab_style(cell_text(weight = "bold"),
			  list(cells_column_spanners())) %>% 
	tab_style(cell_text(size = "medium"),
			  list(cells_body(),
			  	 cells_stub()))
```

The research reported in this article was conducted in accordance with the principles expressed in the Declaration of Helsinki and approved by the local ethical committee (the clinical research ethical committee of the Parc de la Salut Mar). Before every testing session, caretakers were asked to read and sign an informed consent form, and were given a small gift at the end of it.

@@

## Vocabulary

We collected vocabulary data from participants. In Oxford, participants were sent the online version of the Oxford Communicative Development Inventory [OCDI, @hamilton2000infant] two weeks before each appointment. In Barcelona, participants were sent the online vocabulary inventory [BVQ](https://github.com/gongcastro/bvq) immediately after the appointment, and were given two weeks to fill it. In Oxford, families filled the vocabulary checklist in the testing language, and when the participant was monolingual, also in the second language. In Barcelona, all participants filled the Catalan and the Spanish versions of the vocabulary checklists. All checklists included the words involved in the trial lists. When filling the vocabulary checklist, families checked each word as being *understood*, *understood and produced* or *none*. Each response to either checklist was aggregated to produce several vocabulary measures:

* **Total vocabulary**: total number of words the child knows, summing up both languages. For instance, a child who knows 210 words in Catalan and 100 words in Spanish would have a total vocabulary size of 210 words.
* **L1 vocabulary**: number of words the child knows in the languages of higher exposure (i.e., testing language). For instance, for a child exposed to 75% Catalan who knows 210 words in Catalan and 100 words in Spanish, their L1 vocabulary size would be 210 words.
* **Conceptual vocabulary**: number of lexicalised concepts, that is, the number of concepts for which the child knows a word in at least one of the languages. For example, a Catalan-Spanish bilingual that knows *taula* and *mesa* -- Catalan and Spanish translations of *table*-- would have a conceptual vocabulary size of 1.

Vocabulary sizes of participants whose families failed to complete the vocabulary checklists were imputed, by assigning them the most likely vocabulary size from a distribution of vocabulary sizes of children of similar age ($\pm 1$ month) and language profile (similar exposure to a second language, $\pm 10%$) that completed the same vocabulary checklist previously.

### L1 vocabulary

This figure represents the vocabulary size in the dominant language of participants in the Y-axis and their linguistic profile in the X-axis. Results are presented separately for each age group. The shape of the violins, represents the distribution of the vocabulary sizes. The box-plot inside each violin represents the median, 25th and 75th percentiles, and the vertical lines represent the minimum and maximum values. vocabulary sizes in the dominant language were calculated as the proportion of items each participant was reported to *understand* or *understand and say* in the language of most exposure. for example, for a participant exposed mostly to Catalan, their vocabulary size in the dominant language was calculated as the proportion of words they were reported to understand and/or say in Catalan, from a total of ~200 words.

```{r}
#| label: fig-vocabulary-l1
#| fig-cap: "L1 vocabulary. Percentage of words in L1 reported to be understood from the vocabulary checklist."
#| fig-height: 5
#| fig-width: 11
sample_size <- participants %>% 
	filter(id %in% fit_0$data$id) %>% 
	count(age_group)

sample_size_lp <- participants %>% 
	filter(id %in% fit_0$data$id) %>% 
	count(age_group, lp) %>% 
	mutate(
		age_group = case_when(
			age_group=="21 months" ~ paste0(age_group, " (N = ", sample_size$n[1], ")"),
			age_group=="25 months" ~ paste0(age_group, " (N = ", sample_size$n[2], ")"),
			age_group=="30 months" ~ paste0(age_group, " (N = ", sample_size$n[3], ")")
		)
	)

# vocabulary
participants %>% 
	mutate(
		age_group = case_when(
			age_group=="21 months" ~ paste0(age_group, " (N = ", sample_size$n[1], ")"),
			age_group=="25 months" ~ paste0(age_group, " (N = ", sample_size$n[2], ")"),
			age_group=="30 months" ~ paste0(age_group, " (N = ", sample_size$n[3], ")")
		),
		is_imputed = factor(ifelse(is_imputed, "Imputed", "Observed"),
							levels = c("Observed", "Imputed"))
	) %>% 
	drop_na(lp) %>%
	filter(valid_participant) %>% 
	ggplot(aes(lp,  vocab_size_l1, colour = is_imputed, fill = is_imputed)) +
	facet_wrap(~age_group, scales = "free_x") +
	geom_hline(yintercept = 0.5, colour = "black", linetype = "dotted") +
	stat_slab(colour = "white", fill = "grey", alpha = 0.75,
			  position = position_nudge(x = 0.3)) +
	geom_boxplot(width = 0.15, size = 1, colour = "black",
				 fill = "white", outlier.colour = NA,
				 position = position_nudge(x = 0.3)) +
	geom_jitter(shape = 1, stroke = 1, width = 0.15, alpha = 0.5, size = 2) +
	geom_text(data = sample_size_lp,
			  aes(x = 1.5, y = 0,  label = paste0("N = ", n), group = lp),
			  colour = "black", show.legend = FALSE, 
			  position = position_dodge(width = 2), inherit.aes = FALSE) +
	labs(y = "Receptive vocabulary", 
		 colour = "Imputed", 
		 fill = "Imputed") +
	scale_y_continuous(limits = c(0, 1), labels = percent) +
	theme(legend.position = "top", 
		  legend.title = element_blank(),
		  panel.grid.major.x = element_blank(),
		  axis.title.x = element_blank(),
		  axis.text = element_text(size = 9))
```

### Total vocabulary

```{r}
#| label: fig-vocabulary-total
#| fig-cap: "Total vocabulary"
#| fig-height: 5
#| fig-width: 11
participants %>% 
	mutate(
		age_group = case_when(
			age_group=="21 months" ~ paste0(age_group, " (N = ", sample_size$n[1], ")"),
			age_group=="25 months" ~ paste0(age_group, " (N = ", sample_size$n[2], ")"),
			age_group=="30 months" ~ paste0(age_group, " (N = ", sample_size$n[3], ")")
		),
		is_imputed = factor(ifelse(is_imputed, "Imputed", "Observed"),
							levels = c("Observed", "Imputed"))
	) %>% 
	drop_na(lp) %>%
	filter(valid_participant) %>% 
	ggplot(aes(lp,  vocab_size_total, colour = is_imputed, fill = is_imputed)) +
	facet_wrap(~age_group, scales = "free_x") +
	geom_hline(yintercept = 0.5, colour = "black", linetype = "dotted") +
	stat_slab(colour = "white", fill = "grey", alpha = 0.75,
			  position = position_nudge(x = 0.3)) +
	geom_boxplot(width = 0.15, size = 1, colour = "black",
				 fill = "white", outlier.colour = NA,
				 position = position_nudge(x = 0.3)) +
	geom_jitter(shape = 1, stroke = 1, width = 0.15, alpha = 0.5, size = 2) +
	geom_text(data = sample_size_lp,
			  aes(x = 1.5, y = 0,  label = paste0("N = ", n), group = lp),
			  colour = "black", show.legend = FALSE, 
			  position = position_dodge(width = 2), inherit.aes = FALSE) +
	labs(y = "Receptive vocabulary", 
		 colour = "Imputed", 
		 fill = "Imputed") +
	scale_y_continuous(limits = c(0, 1), labels = percent) +
	theme(legend.position = "top", 
		  legend.title = element_blank(),
		  panel.grid.major.x = element_blank(),
		  axis.title.x = element_blank(),
		  axis.text = element_text(size = 9))
```

### Conceptual vocabulary

```{r}
#| label: fig-vocabulary-conceptual
#| fig-height: 5
#| fig-width: 11
participants %>% 
	mutate(
		age_group = case_when(
			age_group=="21 months" ~ paste0(age_group, " (N = ", sample_size$n[1], ")"),
			age_group=="25 months" ~ paste0(age_group, " (N = ", sample_size$n[2], ")"),
			age_group=="30 months" ~ paste0(age_group, " (N = ", sample_size$n[3], ")")
		),
		is_imputed = factor(ifelse(is_imputed, "Imputed", "Observed"),
							levels = c("Observed", "Imputed"))
	) %>% 
	drop_na(lp) %>%
	filter(valid_participant) %>% 
	ggplot(aes(lp,  vocab_size_conceptual, colour = is_imputed, fill = is_imputed)) +
	facet_wrap(~age_group, scales = "free_x") +
	geom_hline(yintercept = 0.5, colour = "black", linetype = "dotted") +
	stat_slab(colour = "white", fill = "grey", alpha = 0.75,
			  position = position_nudge(x = 0.3)) +
	geom_boxplot(width = 0.15, size = 1, colour = "black",
				 fill = "white", outlier.colour = NA,
				 position = position_nudge(x = 0.3)) +
	geom_jitter(shape = 1, stroke = 1, width = 0.15, alpha = 0.5, size = 2) +
	geom_text(data = sample_size_lp,
			  aes(x = 1.5, y = 0,  label = paste0("N = ", n), group = lp),
			  colour = "black", show.legend = FALSE, 
			  position = position_dodge(width = 2), inherit.aes = FALSE) +
	labs(y = "Receptive vocabulary", 
		 colour = "Imputed", 
		 fill = "Imputed") +
	scale_y_continuous(limits = c(0, 1), labels = percent) +
	theme(legend.position = "top", 
		  legend.title = element_blank(),
		  panel.grid.major.x = element_blank(),
		  axis.title.x = element_blank(),
		  axis.text = element_text(size = 9))
```

## Design

@@

### Data processing

```{r} 
#| label: tbl-attrition-group
#| tbl-cap: "Summary of the dataset."
attrition %>% 
	left_join(select(participants, age_group, lp, id, date_test)) %>% 
	add_count(age_group, lp, name = "n_trials") %>% 
	group_by(id, lp, trial_type, date_test, age_group, valid_participant, n_trials) %>% 
	summarise(valid_trial_count = sum(valid_trial, na.rm = TRUE),
			  .groups = "drop") %>% 
	pivot_wider(names_from = trial_type, values_from = valid_trial_count, values_fill = 0) %>% 
	clean_names() %>% 
	add_count(lp, age_group) %>%
	group_by(lp, age_group, n, n_trials) %>% 
	summarise(across(cognate:unrelated, lst(sum, mean, sd)),
			  n_valid = sum(valid_participant),
			  .groups = "drop") %>%
	relocate(n_valid, .after = n) %>% 
	mutate(n_prop = n_valid/n) %>% 
	arrange(age_group, lp) %>% 
	relocate(age_group, lp, n_valid, n_prop, matches("cognate"), matches("non_cognate"), matches("unrelated")) %>% 
	gt(groupname_col = "age_group", rowname_col = "lp")  %>%
	cols_hide(c(n, n_trials)) %>% 
	fmt_number(matches("_mean|_sd"), decimals = 1) %>%
	fmt_percent(matches("_prop"), decimals = 1) %>% 
	cols_merge_n_pct(col_n = n_valid, col_pct = n_prop, autohide = TRUE)  %>% 
	tab_spanner("Non-cognate", matches("non_cognate")) %>%
	tab_spanner("Cognate", starts_with("cognate")) %>%
	tab_spanner("Unrelated", matches("unrelated")) %>%
	tab_spanner("Related", matches("cognate")) %>%

	cols_merge_uncert(col_val = cognate_mean, col_uncert = cognate_sd) %>% 
	cols_merge_uncert(col_val = non_cognate_mean, col_uncert = non_cognate_sd) %>% 
	cols_merge_uncert(col_val = unrelated_mean, col_uncert = unrelated_sd) %>% 
	cols_label(age_group = "Age",
			   n_valid = "Participants (included)",
			   cognate_sum = "Trials",
			   cognate_mean = "Mean ± SD",
			   non_cognate_sum = "Trials",
			   non_cognate_mean = "Mean ± SD",
			   unrelated_sum = "Trials",
			   unrelated_mean = "Mean ± SD") %>% 
	summary_rows(groups = TRUE,
				 columns = where(is.integer), 
				 fns = list("Sum" = ~sum(.)), 
				 formatter = fmt_number,
				 decimals = 0) %>% 
	summary_rows(groups = TRUE,
				 columns = c(where(is.numeric), -n_valid), 
				 fns = list("Mean" = ~mean(.)),
				 formatter = fmt_number,
				 decimals = 1) %>% 
	grand_summary_rows(columns = where(is.integer),
					   fns = list("N" = ~sum(.)),
					   formatter = fmt_number,
					   decimals = 0) %>% 
	tab_style(cell_text(weight = "bold"),
			  list(cells_column_spanners(),
			  	 cells_grand_summary(),
			  	 cells_row_groups(),
			  	 cells_stub_grand_summary(),
			  	 cells_stub_summary())) 
```

### Predictors

### Model

$$
\begin{aligned}
\textbf{Likelihood:} \\
y_{ij} &\sim \text{Bernoulli}(p)
\end{aligned}
$$ {#eq-likelihood}

where:

- $y$ is an observed response ($y \in \{\text{No, Understands, Understands and Says}\}$)
- $i$ is the participant index
- $j$ is the translation equivalent (TE) index
- $p$ is the probability f target looking ($p \in (0, 1)$)

$p$ is then estimated using a logit regression model as indicated in @eq-linear.

$$
\begin{aligned}
\textbf{Linear model:} \\
logit(p) = \text{ln} \frac{p}{1-p} &= (\beta_{0} + u_{0_{i}}) + \\
& (\beta_{1} + u_{1_{i}}) · \text{Age group}_{i} + & \\
& \beta_{2} · \text{Language Profile}_{i} + & \\
& (\beta_{3} + u_{2_{i}}) · \text{Trial type}_{i} + & \\
& \beta_{4} · (\text{Age group}_{i} \times \text{Language Profile}_{i}) & \\
& (\beta_{5} + u_{3_{i}}) · (\text{Age group}_{i} \times \text{Trial type}_{i}) + & \\
& \beta_{6} · (\text{Language Profile}_{ij} \times \text{Trial type}_{i}) & \\
& \beta_{7} · (\text{Age group}_{i} \times \text{Language Profile}_{i} \times \text{Trial type}_{i}) & \\
\end{aligned}
$$ {#eq-linear}

where:

- $i$ and $j$ index the participant and translation equivalent (TE)
- $\beta_{0_k}$ is the fixed coefficient of the regression model for the intercept of threshold $k$
- $u_{0_{i}}$ and $w_{0_{j}}$ are the by-participant and by-TE adjustments for $\beta_{0_{k}}$ (i.e., random intercepts), respectively
- $\beta_{1-8}$ are the fixed coefficients of the regression model for the predictors of interest
- $u_{1-8_{i}}$ and $w_{1-3_{j}}$ are the by-participant and by-TE adjustments for$\beta_{1-8}$ (i.e., random slopes), respectively

$$
\begin{aligned}
\\
\textbf{Prior:} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.1) & [\mbox{Intercept/response category threshold}] \\
\beta_{1} &\sim \mathcal{N}(1, 0.1) & [\mbox{Age population-level coefficient}]\\
\beta_{2-8} &\sim \mathcal{N}(0, 1) & [\mbox{Rest of population-level coefficients}] \\
u_{0-8_{i}} &\sim \mathcal{N}(0, \sigma_{u_{0-8_{i}}}) & [\mbox{Participant-level coefficient variability}] \\
w_{0-3_{j}} &\sim \mathcal{N}(0, \sigma_{w_{0-3_{j}}}) & [\mbox{TE-level coefficient variability}] \\\\
&&\mbox{[Participant-level coefficient variability]} \\ \\
\Bigg(\begin{smallmatrix}
u_{k_{0}} \\ 
u_{1_{i}} \\ 
\vdots \\ 
u_{8_{i}} 
\end{smallmatrix}\Bigg) &\sim \mathcal{N} 
\Bigg(\Bigg(\begin{smallmatrix}0 \\
0 \\ 
\vdots \\
0\end{smallmatrix}\Bigg), \Sigma_{u}\Bigg) \\
\Sigma_{u} &= \Bigg(\begin{smallmatrix} \\
\rho_{u_{0}} & \rho_{u_{0}} \sigma_{u_{0_{k}}} \sigma_{u_{1}} & \dots & \rho_{u_{0}} \sigma_{u_{0}} \sigma_{w_{8}}\\ 
\rho_{u_{1}} \sigma_{u_{1}} \sigma_{u_{0}} & \rho_{u_{1}} & \dots & \rho_{u_{1}} \sigma_{u_{1}} \sigma_{u_{8}}\\ 
\vdots & \vdots & \vdots & \vdots \\
\rho_{8} \sigma_{u_{8}} \sigma_{u_{0_{k}}} & \dots & \dots & \rho_{u_{8}} \end{smallmatrix}\Bigg) \\
\sigma_{u_{0-8}} &\sim \mathcal{N_{+}}(1, 0.1) \\
\rho_{u} &\sim LKJcorr(2) \\
\\
&&\mbox{[TE-level coefficient variability]} \\ \\
\Bigg(\begin{smallmatrix}
w_{k_{0}}\\ 
w_{1_{j}} \\ 
\vdots \\ 
w_{3_{j}} 
\end{smallmatrix}\Bigg) &\sim \mathcal{N} \Bigg(\Bigg(\begin{smallmatrix}
0\\ 
0 \\ 
\vdots \\
0 
\end{smallmatrix}\Bigg), \Sigma_{w}\Bigg) \\
\Sigma_{w} &= \Bigg(\begin{smallmatrix} \\
\rho_{w_{0}} & \rho_{w_{0}} \sigma_{w_{0_{k}}} \sigma_{w_{1}} & \dots & \rho_{w_{0}} \sigma_{w_{0}} \sigma_{w_{3}}\\ 
\rho_{w_{1}} \sigma_{w_{1}} \sigma_{w_{0}} & \rho_{w_{1}} & \dots & \rho_{w_{1}} \sigma_{w_{1}} \sigma_{w_{3}}\\ 
\vdots & \vdots & \vdots & \vdots \\
\rho_{3} \sigma_{w_{3}} \sigma_{w_{0_{k}}} & \dots & \dots & \rho_{w_{3}} \end{smallmatrix}\Bigg) \\
\sigma_{w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.1) \\
\rho_{w_{0-3}} &\sim LKJcorr(2)
\end{aligned}
$$ {#eq-prior}

```{r}
#| label: fig-prior-epred
#| fig-width: 9
#| fig-height: 6
nd <- tibble(time_bin = seq(0, 20, 0.1)) %>% 
	poly_add_columns(time_bin, degree = 3, prefix = "ot", scale_width = 1)

m <- add_epred_draws(nd, fit_prior, ndraws = 50, re_formula = NA) %>% 
	mutate(.epred = plogis(.epred))

ggplot(m, aes(time_bin, .epred)) +
	geom_hline(yintercept = 0.5, colour = "grey", size = 1) +
	geom_line(aes(group = .draw), colour = clrs[4], size = 1, alpha = 0.5) +
	labs(x = "Time (ms)",
		 y = "Proportion of target looking time") +
	scale_y_continuous(labels = percent) +
	scale_x_continuous(labels = function(x) format(x*100, big.mark = ","))
```

@@

## Attrition

::: {.tabset-panel}

### By participants

```{r}
#| label: fig-attrition-participants
#| fig-height: 4
n_text <- participants %>% 
	left_join(distinct(attrition, id, valid_participant)) %>%
	left_join(vocabulary) %>% 
	drop_na(valid_participant) %>% 
	mutate(
		valid_status = case_when(
			valid_participant & !is_imputed ~ "Valid (observed CDI)",
			valid_participant & is_imputed ~ "Valid (imputed CDI)",
			!valid_participant ~ "Excluded"
		)
	) %>% 
	filter(valid_status != "Excluded") %>% 
	count(lp, age_group) %>% 
	mutate(n_label = paste0(n, " valid"),
		   valid_status = "Valid (observed CDI)")

participants %>% 
	left_join(distinct(attrition, id, valid_participant)) %>%
	left_join(vocabulary) %>% 
	drop_na(valid_participant) %>% 
	mutate(valid_status = case_when(
		valid_participant & !is_imputed ~ "valid_cdi_observed",
		valid_participant & is_imputed ~ "valid_cdi_imputed",
		!valid_participant ~ "excluded")) %>% 
	add_count(age_group, lp, name = "n_total") %>% 
	count(age_group, lp, n_total, valid_status) %>% 
	pivot_wider(names_from = valid_status, values_from = n, values_fill = 0) %>% 
	relocate(age_group, lp, n_total, valid_cdi_observed, valid_cdi_imputed, excluded) %>% 
	mutate(across(matches("valid_|excluded"), ~./n_total, .names = "{.col}_prop")) %>% 
	gt(rowname_col = "lp", groupname_col = "age_group") %>% 
	fmt_percent(matches("prop"), decimals = 1) %>% 
	cols_merge_n_pct(col_n = valid_cdi_observed, col_pct = valid_cdi_observed_prop) %>% 
	cols_merge_n_pct(col_n = valid_cdi_imputed, col_pct = valid_cdi_imputed_prop) %>% 
	cols_merge_n_pct(col_n = excluded, col_pct = excluded_prop) %>% 
	tab_spanner("Included participants", matches("valid")) %>% 
	tab_spanner("Excluded participants", matches("excluded")) %>% 
	cols_label(n_total = "N",
			   valid_cdi_observed = "Collected CDI",
			   valid_cdi_imputed = "Missing CDI",
			   excluded = "") %>% 
	summary_rows(groups = TRUE, columns = where(is.integer), 
				 fns = list("N" = ~sum(.)),
				 formatter = fmt_number,
				 decimals = 0) %>% 
	grand_summary_rows(columns = where(is.integer), 
					   fns = list("N" = ~sum(.)),
					   formatter = fmt_number,
					   decimals = 0) %>% 
	tab_style(cell_text(weight = "bold"),
			  list(cells_column_spanners(),
			  	 cells_grand_summary(),
			  	 cells_row_groups(),
			  	 cells_stub_grand_summary(),
			  	 cells_stub_summary()))
```

### By stimuli properties

```{r}
#| label: tbl-attrition-predictors

attrition %>% 
	left_join(distinct(participants, id, age_group, test_language, list, version, list)) %>% 
	left_join(distinct(stimuli, test_language, version, list, trial_id, trial_type,
					   freq_target, freq_prime, familiarity_prime, familiarity_target)) %>% 
	mutate(valid_status = ifelse(valid_status != "Valid",  "Excluded", valid_status)) %>% 
	drop_na() %>% 
	group_by(trial_type, valid_status, age_group) %>% 
	summarise(n = n(),
			  across(matches("freq_|familiarity_"), lst(mean, sd)),
			  .groups = "drop") %>% 
	pivot_wider(names_from = valid_status,
				values_from = c(n, matches("freq_|familiarity_")),
				values_fill = 0) %>%
	clean_names() %>% 
	relocate(age_group, trial_type, n_valid, n_excluded, 
			 matches("_valid")) %>%

	gt(rowname_col = "trial_type", groupname_col = "age_group") %>% 
	fmt_number(matches("n_"), decimals = 0) %>% 
	fmt_number(matches("freq_"), decimals = 1) %>% 
	fmt_number(matches("familiarity"), scale_by = 100, decimals = 1) %>%

	tab_spanner("N", c(n_valid, n_excluded)) %>%

	tab_spanner("Prime", id = 1, c(freq_prime_mean_valid, freq_prime_mean_excluded)) %>%
	tab_spanner("Target", id = 2, c(freq_target_mean_valid, freq_target_mean_excluded)) %>% 
	tab_spanner("Prime", id = 3, c(familiarity_prime_mean_valid, familiarity_prime_mean_excluded)) %>% 
	tab_spanner("Target", id = 4, c(familiarity_target_mean_valid, familiarity_target_mean_excluded)) %>%
	tab_spanner("Frequency", starts_with("freq_")) %>% 
	tab_spanner("Familiarity (%)", starts_with("familiarity_")) %>%

	cols_merge_uncert(col_val = freq_prime_mean_valid, col_uncert = freq_prime_sd_valid) %>% 
	cols_merge_uncert(col_val = freq_prime_mean_excluded, col_uncert = freq_prime_sd_excluded) %>% 
	cols_merge_uncert(col_val = freq_target_mean_valid, col_uncert = freq_target_sd_valid) %>% 
	cols_merge_uncert(col_val = freq_target_mean_excluded, col_uncert = freq_target_sd_excluded) %>% 
	cols_merge_uncert(col_val = familiarity_prime_mean_valid, col_uncert = familiarity_prime_sd_valid) %>% 
	cols_merge_uncert(col_val = familiarity_prime_mean_excluded, col_uncert = familiarity_prime_sd_excluded) %>% 
	cols_merge_uncert(col_val = familiarity_target_mean_valid, col_uncert = familiarity_target_sd_valid) %>%
	cols_merge_uncert(col_val = familiarity_target_mean_excluded, col_uncert = familiarity_target_sd_excluded) %>%

	cols_label(n_excluded = "Exc.",
			   n_valid = "Inc.",
			   freq_prime_mean_valid = "Inc.",
			   freq_prime_mean_excluded = "Exc.",
			   freq_target_mean_valid = "Inc.",
			   freq_target_mean_excluded = "Exc.",
			   familiarity_prime_mean_valid = "Inc.",
			   familiarity_prime_mean_excluded = "Exc.",
			   familiarity_target_mean_valid = "Inc.",
			   familiarity_target_mean_excluded = "Exc.") %>% 
	summary_rows(groups = TRUE,
				 columns = matches("freq_") ,
				 fns = list("Mean" = ~mean(.)), 
				 formatter = fmt_number,
				 decimals = 1) %>% 
	summary_rows(groups = TRUE,
				 columns = matches("familiarity_"), 
				 fns = list("Mean" = ~mean(.)),
				 formatter = fmt_number,
				 scale_by = 100,
				 decimals = 1) %>% 
	grand_summary_rows(columns = c(n_valid, n_excluded),
					   fns = list("N" = ~sum(.)),
					   formatter = fmt_number,
					   decimals = 0) %>% 
	grand_summary_rows(columns = matches("familiarity_"),
					   fns = list("Mean" = ~mean(.)),
					   formatter = fmt_number,
					   scale_by = 100,
					   decimals = 1) %>% 
	grand_summary_rows(columns = matches("freq_"),
					   fns = list("Mean" = ~mean(.)),
					   formatter = fmt_number,
					   decimals = 1) %>% 
	tab_style(cell_text(align = "center"),
			  cells_column_labels()) %>% 
	tab_style(cell_text(weight = "bold"),
			  list(cells_column_spanners(),
			  	 cells_grand_summary(),
			  	 cells_row_groups(),
			  	 cells_stub_grand_summary(),
			  	 cells_stub_summary())) %>% 
	tab_footnote("Included trials.", cells_column_labels(matches("_valid"))) %>% 
	tab_footnote("Excluded trials.", cells_column_labels(matches("_excluded")))
```

>>>>

---
title: "Developmental trajectories of word recognition: a priming study in bilinguals"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - ref: upf
    corresponding: true
  - name: Serene Siow
    affiliations:
      - ref: oxf
  - name: Núria Sebastian-Galles
    affiliations:
      - ref: upf
  - name: Kim Plunkett
    affiliations:
      - ref: oxf
affiliations:
  - id: upf
    name: Center for Brain and Cognition, Universitat Pompeu Fabra
  - id: oxf
    name: Department of Experimental Psychology, University of Oxford

pagetitle: Cross-linguistic similarity and  word acquisition
abstract: "Previous studies have provided evidence that lexical access is language non-selective in bilinguals: recognising and producing words in one language activates lexical representations of words in the other language (e.g., Costa et al., 2000; Thierry & Wu, 2007). It has been suggested that this parallel activation is already present during toddlerhood (Jardak & Byers-Heinlein, 2019; Von Holzen & Mani, 2012), but it is unclear how it impacts the developing lexicon. In the present study, we tested Catalan-Spanish and English-Spanish bilinguals in a word recognition task, in which they were presented with words in their dominant language exclusively. In each trial, two pictures were presented side-by-side and one of them was named (target picture). Participants' visual preference for the named picture was taken as an indicator of word recognition. Each pair of pictures was preceded by the presentation of a silent image (prime picture). We designed three types of trials: (1) cognate trials in which prime and target labels shared phonological onset and the prime label was a cognate (e.g., flower-flor / fork), (2) non-cognate trials, in which prime and target labels shared phonological onset but the prime label was a non-cognate (e.g., frog-rana / fork), and (3) unrelated trials, in which prime and target labels did not share onset (unrelated trials; car-coche / fork). In line with previous studies, we expected participants to generate implicit labels for the prime pictures, which should interfere with target word recognition when both words share phonological onset (e.g., Mani & Plunkett, 2011). Additionally, under the hypothesis that bilingual participants would activate labels for the prime pictures in both languages, we predicted that interference would be stronger after cognate primes (labels from both languages share phonological overlap with target word recognition) than after non-cognate primes (only the label in the target language share phonological overlap with target word recognition). We tested bilinguals (n = 46) and same-aged monolingual controls, n = 123) at three age points (21, 25, and 30 months) to investigate how any cross-language priming effect emerged or changed across these ages."
thanks: "This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We thank Xavier Mayoral, Silvia Blanch, Cristina Cuadrado, and Irina Lepadatu for their technical support. We also thank Cristina Dominguez, Katia Pistrin, and Nicola Gillen for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances."
keywords: word recognition, cognate, vocabulary, bilingualism, eye-tracking, growth curve analysis, bayesian
csl: assets/bibliography/apa7.csl
bibliography: assets/bibliography/references.bib
format:
  pdf:
    pdf-engine: lualatex
    number-sections: true
    whitespace: small
  docx:
  	reference-doc: assets/templates/template.docx
linestretch: 1.5
toc: false
fig-dpi: 1000
echo: false
warning: false
---

@@

# Introduction

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/ckydb/)) and GitHub ([https://github.com/gongcastro/cognate-priming](https://github.com/gongcastro/cognate-priming)) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

@@

@@

## Procedure

## Apparatus

@@

@@

# Discussion
